Astrophysical calibration of gravitational-wave detectors

### 2016-02-03 docs/response_to_referee.txt

We would like to thank the referee for their very useful comments.
We have included response to them inline below and make note where this
has resulted in changes to the manuscript. These changes are displayed in red
in the resubmitted version of the manuscript.

@@

Major comment: "In section 5, I have a problem with the assumption of the sGRB jet
opening angle being modelled as a half-normal distribution. Figure 3
in Fong et al. (2014) shows that the long GRB distribution may be
modelled by something like a Rayleigh distribution (or something
similar to give a tail to high values), but with little weight at
small angles. If the same is true of the sGRB population (and from
Figure 3 the long tail seems to be present here too) then a
half-normal distribution for the opening angle is not a good match to
the data. Did you try other priors on the inclination angle which may
better represent the true distribution? I ask that you try to use a
better prior for the inclination angle, or alternatively provide me
with a better justification of your current choice."

Response: Our poor description regarding how we use the opening angle (theta_j)
distribution to constrain the distribution on the orientation angle (iota) of
the systems has caused confusion here. We are not really interested
in the jet opening angle distribution, but are just trying to somehow use
it as a proxy for getting an idea of the orientation angle distribution,
which would not be expected to look the same. Under the assumption that to
observe a sGRB we must have theta_j/2 >= |iota|, and assuming a Rayleigh
distribution on theta_j, the distribution on |iota| does become a half-normal
distribution. In fact our choice of a half-normal for |iota| with a sigma value of
14.8 degs is rather conservative* (in that it extends to larger iota values
[this is something we would plan to correct in future studies]) compared to the
half-normal distribution that you truly get based a Rayleigh distribution in 
theta_j with a median of 10 degs. We have added text to the manuscipt to
hopefully make this clearer. If the referee would like to see more detail on
these we can send them further information.

*this conservative nature is in part due to one of us (M Pitkin) mistakenly
assuming the opening angle referred to the half width of the beam rather than
the full beam width.

@@

Minor comments
--------------

Comment: "The first sentence should be followed by a reference to Abadie et
al. (2010), or a similar updated article on predicted detection rates
if available."

Response: We agree and have added the Abadie et al. (2010) reference here.

@@

Comment: ""error in the amplitude calibration" -> is this the strain
amplitude or the error signal output. Clarify either way."

Response:

Comment: ""within a control feedback loop (see [7]), which are subject to
uncertainties." -> This is quite a vague sentence. What kind of
uncertainties?"

Response: This has been changed to "...within a control feedback loop (see [7]),
each of which are subject to measurement uncertainties."

@@

Comment: ""larger amplitude (e.g. be closer)" -> it would only appear to be
closer if we had a handle on the other parameters affecting the
amplitude e.g. the chirp mass. We obviously get the latter from the
signal's phase evolution, but this point should be clarified."

Response:

Comment: "How realistic is the assumption of a constant calibration scale
factor over the full LIGO band? What physical mechanism would lead us
to make a poor estimation of the length response function, and why
would other calibration tests fail to mitigate the effect dealt with
in this article?"

Response: Our analysis uses the frequency range from 20-400Hz, so our assumption
of a constant calibration scale factor only relates to this band rather than the
full LIGO bandwidth. However, having just one scale factor should be a
conservative case and in the future we would like to look at how having
some frequency dependence to the calibration scale factor would effects results.

We do not consider what physical effects could cause a poor estimation of the
length response function. We mainly put our proposal forward as an independent
way of estimating the length response function.

@@

Comment: ""uniform distribution on the sky" -> does this mean uniform in
right ascension and uniform in cosine(90-declination)?"

Response:

Comment: "Your range of injected calibration scale factors is quite narrow,
meaning you only really tested situations where the factor was pretty
close to optimal. Did you test cases where the detectors were very
poorly calibrated? If not, then I highly suggest that you do so, or
adequately comment on expectations from these tests."

Response: We have tested situations were the detectors were very poorly
calibrated e.g. scale factors ranging from ~0.1 to ~10. Even for these
very poorly calibrated cases the widths of the scale factor distributions
stay consistent with cases were the scale factor is close to 1. We have
added some additional text to footnote 7 about this.

@@

Comment: ""calculated by averaging 64 separate noisy realisations of the PSD
and scaled with the same calibraion scale factor as applied to the
injection and noise" -> firstly the typo in "calibration" should be
corrected. Also, does this mean that the calibration scale factor
appears in "S" as a parameter also during your MCMC? If so, why does
it not explicitly appear in the denominator of the exp[] of equation
6?"

Response: The typo has been corrected. The calibration scale factor does
not appear in "S" during the MCMC or the denominator of equation 6. The
description in this section is saying how the simulated data was created,
whereas during the MCMC it is the only the expect signal waveform
that has the scale factor applied to it to try and match the data (to which
the calibration factor has already been applied during its creation).

Comment: "In the paragraph on page 6 beginning "In figure 3..." there is some
explanation of why the H1 and L1 uncertainties saturate out to ~25%.
The explanation did not seem very clear to me, and does not address
why V1 does not show a similar saturation. What is the significance of
the value of ~25%? Is there some influence of the finiteness of the
detected population at larger distances that is influencing the width
of the uncertainty distribution?"

Response: We have added the additional text stating "the weakest signals that could still
be detected would always have a single detector \ac{SNR} of $\sim 5.5$ no matter
their distance, hence the plateau.

The issue of the V1 result not saturating is addressed in the manuscript in the lines
"The largest SNR contribution will generally come from the two aLIGO detectors and thus
the detection criteria (SNR threshold) will not apply to the AdV result. Hence, the SNR
in AdV can be small and thus the ability to constrain its scale factor becomes poor (although it
still provides information that the calibration is not grossly inaccurate)."

@@

Comment: "In figure 1 the tick labels for the chirp mass are overlaid with
the label. Better labelling or figure formatting is needed here, as
well as units where appropriate. I like the different edge coloring
for the relevant scale factor boxes, but perhaps these red edges could
be thickened for black and white printing."

Response:

Comment: "No mention is given to the number of MCMC iterations, the length of
burn-in, or how convergence was ascertained. A brief comment would be
appropriate."

Response: As a proxy to check for convergence of the chains we check that the
scale parameter posterior histograms do not contain many disjoint modes. We have
added this as a footnote into the paper. The actual tests we use are that 90% or
more of the histogram bins are non-zero and the standard deviation
of the posterior is less than 1/6th of the total histogram range.

### 2016-02-03 docs/cosmic_cal.tex
Most of my comments have been addressed in the response and manuscript

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of \acp{CBC} per year. One
such class of these cataclysmic events, the inspiral and merger of \ac{BNS}
systems will be detected out to a maximum range of $\approx 450$ Mpc. Assuming
the current best estimates for the cosmological parameters, this is equivalent
to a redshift $z\approx 0.1$. As noted by~\citet{1986Natur.323..310S}, \ac{CBC}
systems can be used as cosmological distance markers, otherwise known as
``standard sirens'' (analogous to \ac{EM} standard candles). The nature of
these sources will allow us to estimate, amongst other parameters, their
luminosity distance without the need for calibration against the cosmic
distance ladder.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer; 
where an elaborate feedback system is used to sustain a defined measurement in arm length 
difference between the moving mirrors. A comprehensive description of the calibration procedure (in 
particular for the LIGO detectors during their fifth science run) can be found in 
\cite{2010NIMPA.624..223A}, and a brief description is given in \cite{Vitale:2012} and references 
therein. For \ac{aLIGO} and \ac{AdV}, at the frequency range used in this project, the error in the 
amplitude calibration is expected to be roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f, t)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that at a particular time we have
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the same scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency (and is real, so has no phase component), but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large
number (several hundred up to $O(1000)$) of simulations at each distance value.
When proposing injections for each simulation the source parameters were
randomly drawn from the prior distributions described in
Section~\ref{sec:priors}, with the sky position being drawn randomly from a
uniform distribution on the sky, the coalescence times held fixed within the
centre of the prior window, and with the calibration scaling factors for each
detector drawn from a Gaussian distribution with a mean of one and standard
deviation of $0.125$ (equivalent to a mean fractional offset of 10\%).  Note
that this distribution is broadly consistent with the scale of calibration
uncertainties that are expected via experimental methods, but is not the same as
the prior that we assume on $\mathcal{C}$ (see Eq.~\ref{eq:scale_prior}) when recovering 
signals\footnote{The reason behind this is that we wanted to use a fairly na\"{i}ve and
conservative prior for signal recovery that had little impact in the results, i.e.\
to produce results domintaed by the likelihood for these parameters.}.

@@

% a discussion of the PSD estimate
For each of the simulated sources we have used the {\tt emcee} python \ac{MCMC}
package~\cite{2013PASP..125..306F} to perform parameter estimation over the
unknown source parameters using the priors as discussed in
Section~\ref{sec:priors}. In each case when calculating the likelihood we used
an estimate of the noise \ac{PSD} based on the advanced detector design
sensitivities (using those given in~\cite{2013arXiv1304.0670L}), but calculated by averaging 
64 separate noisy realisations of the \ac{PSD} and scaled
with the same \refresp{calibration} scale factor as applied to the injection and noise\footnote{We do not account for there 
being a potential difference
between the estimated \ac{PSD} and the actual \ac{PSD} of the analysed section
of data as described in e.g.\ \cite{2013PhRvD..88h4044L}. This difference would
be very highly correlated with the calibration scale factor, so in reality our
estimate of the scale factor would be a combination of the calibration offset
and the difference in the \ac{PSD}. As such our results on real data would be
an upper limit on the calibration scale factor.}. This has provided posterior probability
distributions on the calibration scale factors for each detector.  Examples of
the marginalised posterior probabilities for a \ac{BNS} system and a \ac{NSBH}
system observed with the three detector network are shown in
figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

@@

% discuss the BNS plots - general behaviour
In figure~\ref{fig:bnsresults} we see that on average the scale factor can be
recovered to equivalent precision in both \ac{aLIGO} detectors, as would be
expected, with uncertainties generally within the 10\% range for sources at
100\,Mpc. This is comparable to previous estimates of the calibration error in
the initial LIGO detectors. An interesting feature is that for distances
$\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a
maximum at $\sim 25\%$, whilst the width of boxes narrows. This is due to our
``detectibility'' criteria, where for all distances we will only be seeing
those sources with a high enough \ac{SNR} that we would consider them
detectable. There will therefore be fewer sources with \ac{SNR} higher than
this criteria at large distances, giving us a narrower range, and we
automatically exclude those with \ac{SNR} that are too small thus truncating
our uncertainty distribution at the upper end.  However, this does show that on
average for sources that are detectable out to 450\,Mpc we would be able to
constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors
to $\sim 20\%$.

>>>>

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of \acp{CBC} per year \refresp{\cite{2010CQGra..27q3001A}}. One
such class of these cataclysmic events, the inspiral and merger of \ac{BNS}
systems will be detected out to a maximum range of $\approx 450$ Mpc. Assuming
the current best estimates for the cosmological parameters, this is equivalent
to a redshift $z\approx 0.1$. As noted by~\citet{1986Natur.323..310S}, \ac{CBC}
systems can be used as cosmological distance markers, otherwise known as
``standard sirens'' (analogous to \ac{EM} standard candles). The nature of
these sources will allow us to estimate, amongst other parameters, their
luminosity distance without the need for calibration against the cosmic
distance ladder.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer\refresp{,}
where an elaborate feedback system is used to sustain a defined measurement in arm length 
difference between the moving mirrors. A comprehensive description of the calibration procedure (in 
particular for the LIGO detectors during their fifth science run) can be found in 
\cite{2010NIMPA.624..223A}, and a brief description is given in \cite{Vitale:2012} and references 
therein. For \ac{aLIGO} and \ac{AdV}, at the frequency range used in this project, the error in the 
amplitude calibration is expected to be roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f, t)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), \refresp{each of which are subject to measurement uncertainties}. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that at a particular time we have
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain \refresp{(i.e.\
the measured length response function $R(f)$ is related to the true length response function $R_{\rm true}(f)$
via $R(f) = \scf R_{\rm true}(f)$)}. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the same scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency (and is real, so has no phase component), but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large
number (several hundred up to $O(1000)$) of simulations at each distance value.
When proposing injections for each simulation the source parameters were
randomly drawn from the prior distributions described in
Section~\ref{sec:priors}, with the sky position being drawn randomly from a
uniform distribution on the sky, the coalescence times held fixed within the
centre of the prior window, and with the calibration scaling factors for each
detector drawn from a Gaussian distribution with a mean of one and standard
deviation of $0.125$ (equivalent to a mean fractional offset of 10\%).  Note
that this distribution is broadly consistent with the scale of calibration
uncertainties that are expected via experimental methods, but is not the same as
the prior that we assume on $\scf$ (see Eq.~\ref{eq:scale_prior}) when recovering 
signals\footnote{The reason behind this is that we wanted to use a fairly na\"{i}ve and
conservative prior for signal recovery that had little impact in the results, i.e.\
to produce results domintaed by the likelihood for these parameters. \refresp{However,
limited studies using $\scf$ values within the much larger range of
$\sim 0.1$ to 10 show distributions with consistent relative widths to
those we find in our main study.}}.

@@

% a discussion of the PSD estimate
For each of the simulated sources we have used the {\tt emcee} python \ac{MCMC}
package~\cite{2013PASP..125..306F} to perform parameter estimation over the
unknown source parameters using the priors as discussed in
Section~\ref{sec:priors}. In each case when calculating the likelihood we used
an estimate of the noise \ac{PSD} based on the advanced detector design
sensitivities (using those given in~\cite{2013arXiv1304.0670L}), but calculated by averaging 
64 separate noisy realisations of the \ac{PSD} and scaled
with the same \refresp{calibration} scale factor as applied to the injection and noise\footnote{We do not account for there 
being a potential difference
between the estimated \ac{PSD} and the actual \ac{PSD} of the analysed section
of data as described in e.g.\ \cite{2013PhRvD..88h4044L}. This difference would
be very highly correlated with the calibration scale factor, so in reality our
estimate of the scale factor would be a combination of the calibration offset
and the difference in the \ac{PSD}. As such our results on real data would be
an upper limit on the calibration scale factor.}. This has provided posterior probability
distributions\refresp{\footnote{\refresp{As a proxy to check for convergence of the MCMC chains
we check that the calibration scale factor posterior histograms do not contain many disjoint modes.}}}
on the calibration scale factors for each detector.  Examples of
the marginalised posterior probabilities for a \ac{BNS} system and a \ac{NSBH}
system observed with the three detector network are shown in
figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

@@

% discuss the BNS plots - general behaviour
In figure~\ref{fig:bnsresults} we see that on average the scale factor can be
recovered to equivalent precision in both \ac{aLIGO} detectors, as would be
expected, with uncertainties generally within the 10\% range for sources at
100\,Mpc. This is comparable to previous estimates of the calibration error in
the initial LIGO detectors. An interesting feature is that for distances
$\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a
maximum at $\sim 25\%$, whilst the width of boxes narrows. This is due to our
``detectibility'' criteria, where for all distances we will only be seeing
those sources with a high enough \ac{SNR} that we would consider them
detectable\refresp{, i.e.\ the weakest signals that could still be detected would
always have a single detector \ac{SNR} of $\sim 5.5$ no matter their distance, hence the
plateau}. \refresp{In addition there will also} be fewer sources with \ac{SNR} higher than
this criteria at large distances, giving us a narrower range, and we
automatically exclude those with \ac{SNR} that are too small thus truncating
our uncertainty distribution at the upper end.  However, this does show that on
average for sources that are detectable out to 450\,Mpc we would be able to
constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors
to $\sim 20\%$.
